Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-17 papers/clusters.tex
import from Rix

\title{Chemical tagging: \\
       Identification of stellar phase-space structures \\
       purely by chemical-abundance similarity}
\author{David~W.~Hogg\altaffilmark{1,2,3,4},
        Andrew~R.~Casey\altaffilmark{5},
        M.~Ness\altaffilmark{4},
        Hans-Walter~Rix\altaffilmark{4},
    and~Daniel~Foreman-Mackey\altaffilmark{6}}
\altaffiltext{1}{SCDA}
\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Phyics,
  New York University, 4 Washington Pl., room 424, New York, NY, 10003, USA}
\altaffiltext{3}{Center for Data Science, New York University, 726 Broadway, 7th Floor, New York, NY 10003, USA}
\altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{5}{IoA}
\altaffiltext{6}{UW}
\email{david.hogg@nyu.edu}

@@

If every star in the Milky Way was born in a molecular cloud that has
unique chemical-abundance characteristics (as a function of time), and
if every star preserves its surface abundances (or significant aspects
of them) over its lifetime, then stars with common birthplaces and
time ought to be identifiable in their surface abundances.
That is, it ought to be possible to use stellar abundances to link up
co-eval stars---even if at the present day they are no longer close to
one another in phase space---and infer a detailed star-formation and
dynamical history for the Galaxy.
This idea has been dubbed ``chemical tagging''.
It was proposed long ago \citep{freeman} and is one of the
principal motivations for a number of surveys, most notably
\apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
In order to measure chemical tags and infer the Galaxy's dynamical
properties and formation history, these surveys are each measuring high resolution,
high signal-to-noise spectra for hundreds of thousands of stars across
the Galaxy's disk, bulge and halo.

Chemical tagging holds the promise of revealing not just the
star-formation history of the Galaxy, but also the accretion history
(as things that fall in are expected to be chemically distinct from
those that form in the parent body; \citealt{something}) and
stellar-orbit diffusion process like radial mixing and radial
migration (for example, \citealt{Roskar2009, quillen}).  After
stars are born---or after a star cluster is accreted and
disrupted---associations or groups will disperse, through either
coherent orbital evolution or else incoherent diffusion, but the
chemical-abundance tags can in principle make the former members
identifiable as co-eval stellar siblings.

Although undeniably promising---and although it has motivated the
launch of expensive, large-scale spectroscopic surveys---the
chemical-tagging idea has yet to yield scientific fruit.
Part of the reason is that the level of chemical specificity required
is very high:
If there are thousands of (relevant) molecular clouds forming stars in
the recent history of the Milky Way, it will take many abundances,
each measured accurately, to make a sufficiently unique fingerprint
that stellar sibling identification will be reliable.

Stellar spectroscopic surveys now have the resolution, signal-to-noise,
wavelength coverage, and sample sizes to deliver chemical tags.
There are, however, two big issues.
The first is that the physical assumptions behind the idea may require
refinement:
There may be chemical-abundance overlaps among open clusters
\citep{blancocuaresma}; coeval groups of stars may have similar tags
but different birth places \citep{mitschang}, and the
chemical-abundance space might be low in dimensionality \citep{what}.
On the other hand, precise studies of stellar twins \citep{melendez, jofre}
and open clusters \citep{bovy} indicates that chemical abundances ought to be
very precisely similar for co-eval stars.

The second issue for chemical tagging---and the one we address
here---is measurement precision (and accuracy).
The current precision on abundance measurements is not high enough
(for example, \citealt{martel, ting}).
That said, it is clear that the \emph{data} are precise enough;
there is enough signal-to-noise at the relevant locations in spectrum
space to deliver high-precision tags.
Our view is that the traditional methods for making abundance
measurements just aren't mature enough:
There are issues with the physical models, and the data-analysis
methods do not make use of all of the information in the data sets.
Improving the models and exploiting the entire information
content in the data is critical if we are going to deliver useful
chemical tags.

Our contribution in this space has been to develop
\thecannon\ \citep{thecannon, ages}, which is a data-driven model for
stellar spectra.
This model can deliver stellar parameters and chemical abundances to
stars, making use of every pixel of every stellar spectrum (that is,
all the information in the data) but making no use at all of physical
models of stars.
It relies only on there being training data---\emph{some} stars for
which parameters and abundances are known and believed.
In companion papers \citep{casey16, ness16} we show that
\thecannon\ can deliver 15 to 19 abundances for stars in the
\apogee\ survey, at precisions higher than even the training data on
which the model is trained.
We say more about this below.
We will show here that \thecannon\ improves chemical abundance measurements
to the point that \emph{chemical tagging is now possible.}
We have thereby effectively saved the collective asses of all the
collaborators in all of the above-referenced large spectroscopic
surveys.

@@

In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
which we can measure 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni).
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---explores the thin disk.
However, it has the disadvantage that it's spatial coverage is swiss
cheese, which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage.

One final note:
We think of this \documentname\ as performing a hypothesis test.
We know that the stellar members of open and globular clusters---stars
that are identified by being close in phase space---contain highly
informative abundance information that identifies them also in
chemical-abundance space.
Does this work the other way around?
Are groups of stars that are identified as being close in abundance
space---collections of stars with near-identical chemical tags---also
found to be close in phase space?
If the answer to this question is ``yes'' then chemical tagging is
\emph{on.}

\section{Sample and abundances}

The details about how we select, reduce, and analyze the \apogee\ data
are given in full detail in the companion papers \citep{casey16, ness16}.
We summarize briefly here.
For the spectroscopic data, we use the \apogee\ \acronym{DR12}
(\cite{dr12}) spectroscopic data reductions for red-giant stars
($\logg < 3.9$) that are not flagged as bad in the
\apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
There are \totalnumber\ stars that make our $\logg$ and flag cuts.

The \apogee\ pipelines deliver stacked (co-added),
pseudo-continuum-normalized spectra, but we do not use these; we
perform our own continuum normalization on individual-exposure data
and co-add it ourselves.
This is required, because the performance of \thecannon\ depends
critically on the methods used for spectral normalization.  It does
not require accurate continuum estimation, but it requires that the
normalization vector be obtained by a \emph{linear operation on the
  data}.
The linear operation we use is a smooth-function fit to pixels
identified as continuum in our previous work \citep{thecannon}.

For the training data labels (used to train \thecannon; see below), we
use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
15-element abundances, but only for the subset of ARC:XXXXX red-giant
stars that have spectral signal-to-noise ratio measurements greater
than 200.
This is a small fraction of the total sample, but these high
signal-to-noise stars are observed to obtain consistent, low-scatter,
sensible abundance-space measurements.

As described in the original paper \citep{thecannon} and the companion
papers \citep{casey16, ness16}, \thecannon\ is operated in two stages.
In the first stage---the traning step---the training set of stars with
measured spectra and known labels (known parameters and abundances) is
used to establish a likelihood function for a set of coefficients that
parameterize a smooth, simple prediction of spectral pixels given
labels.
\thecannon\ finds the set of coefficients that optimizes the
training-step likelihood function.
In the implementation used in this \documentname, the spectral
prediction is quadratic in labels (including all cross-terms to second
order), and there are 17 labels ($\teff$, $\logg$, and 15 abundances
referenced to H) and 8500 spectral pixels, so there are on the order
of 1.4 million coefficients set by a convex optimization in the
training step.

In the second stage---the test step---each unlabeled star is used to
establish a single-star likelihood function for labels, holding the
coefficients fixed.
\thecannon\ finds the labels that minimize the single-star likelihood
function for each test-set star.
This optimization is not convex, but it is embarrassingly parallel.
For the results shown here, the training step takes hours and the test
step takes less than an hour on a small research cluster in Cambridge.

>>>>

\title{Chemical tagging can work: \\
       Identification of stellar phase-space structures \\
       purely by chemical-abundance similarity}
\author{David~W.~Hogg\altaffilmark{1,2,3,4},
        Andrew~R.~Casey\altaffilmark{5},
        M.~Ness\altaffilmark{4},
        Hans-Walter~Rix\altaffilmark{4},
    and~Daniel~Foreman-Mackey\altaffilmark{6}}
\altaffiltext{1}{SCDA}
\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Phyics,
  New York University, 4 Washington Pl., room 424, New York, NY, 10003, USA}
\altaffiltext{3}{Center for Data Science, New York University, 726 Broadway, 7th Floor, New York, NY 10003, USA}
\altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{5}{IoA}
\altaffiltext{6}{UW}
\email{david.hogg@nyu.edu}

@@

Ensembles of stars in the Milky Way are born in molecular clouds,
which are presumed to be near-homogeneous in their chemical element composition.
However, most stars are born in unbound associations, or are born in star clusters that disperse rapidly: they will eventually end up in very different parts of phase space, on differing orbits.
Yet, if every star preserved its photospheric element abundances over its lifetime (at least for most elements), then stars of common birth origin ought to be identifiable through their detailed photospheric abundances, long after any spatial or phase-space proximity has vanished.

This idea, dubbed ``chemical tagging'', was proposed long ago \citep{freeman} and is one of the
principal motivations for a number of surveys, most notably
\apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
In order to determine the precise abundance labels for chemical tagging,  these surveys are each measuring high resolution,
high signal-to-noise spectra for hundreds of thousands of stars across
the Galaxy's disk, bulge and halo.

Chemical tagging holds the promise of revealing not just the
star-formation history of the Galaxy, but also the accretion history
(as things that fall in are expected to be chemically distinct from
those that form in the parent body; \citealt{something}) and
stellar-orbit diffusion process like radial mixing and radial
migration (for example, \citealt{Roskar2009, quillen}).  After
stars are born---or after a star cluster is accreted and
disrupted---associations or groups will disperse, through either
internal two-body processes or external tidal forces.

Although undeniably promising---and motivating the 
launch of costly large-scale spectroscopic surveys---
chemical-tagging as a search technique has yet to be proven in practice: identifying stars of common birth origin {\it purely} on the basis of their near-identical abundances patterns, without any consideration of position or velocity.

Part of the reason is that the level of abundance specificity required
is very high:
If there are thousands of (relevant) molecular clouds forming stars in
the recent history of the Milky Way, clumps of stars can only stand out in a high-dimensional abundance space: therefore accurate, or at least precise, measurements of many different abundances are needed for stellar sibling to have sufficiently unique fingerprints.
Stellar spectroscopic surveys now have the resolution, signal-to-noise,
wavelength coverage, and sample sizes to deliver many different chemical tags for each star.

There are, however, two big issues.
The first is that the physical assumptions behind the idea may require
refinement:
There may be chemical-abundance overlaps among open clusters
\citep{blancocuaresma}; coeval groups of stars may have similar tags
but different birth places \citep{mitschang}, and the
chemical-abundance space might be low in dimensionality \citep{what}.
On the other hand, precise studies of stellar twins \citep{melendez, jofre}
and open clusters \citep{bovy} indicates that chemical abundances ought to be
very precisely similar for co-eval stars.

The second issue for chemical tagging---and the one we address
here---is measurement precision (and accuracy).
The current precision on abundance measurements in the published survey catalogs is not high enough (for example, \citealt{martel, ting}).
However, our recent work \citep{thecannon} suggests that \emph{data} are precise enough: there is enough signal-to-noise at the relevant 
locations in spectrum space to deliver high-precision tags.
The existence of very large data sets, homogeneous in data/spectrum space, suggest that one can use data-driven approaches to determining stellar abundances, which can considerably outperform traditional methods.
These traditional approaches are based on ab-initio physical models
that have shortcomings that become apparent in this age of high-quality spectra; and the data-analysis methods do not make use of all of the information in the data sets.
Improving the models and exploiting the entire information
content in the data is critical if we are going to deliver useful
chemical tags.

Our specific contribution in this space has been to develop
\thecannon\ \citep{thecannon, ages}, which is a data-driven model for
stellar spectra.
This model can deliver stellar parameters and chemical abundances to
stars, making use of every pixel of every stellar spectrum (that is,
all the information in the data) but making no use at all of physical
models of stars.
It relies only on there being training data---\emph{some} stars for
which parameters and abundances are known and believed.
In companion papers \citep{casey16, ness16} we show that
\thecannon\ can deliver 15 to 19 abundances for stars in the
\apogee\ survey, at precisions higher than even the training data on
which the model is trained.
We say more about this below.
We will show here that \thecannon\ improves chemical abundance measurements
to the point that \emph{chemical tagging is now possible.}
We have thereby effectively saved the collective asses of all the
collaborators in all of the above-referenced large spectroscopic
surveys.

@@

In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni), using \thecannon\ \citep{thecannon}.
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---explores the thin disk.
However, it has the disadvantage that it's spatial coverage is swiss
cheese, which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage.

One final note:
we think of this \documentname\ as performing a proof of concept.
We know that the stellar members of open and globular clusters---stars
that are identified by being close in phase space---contain highly
informative abundance information that identifies them also in
chemical-abundance space.
Does this work the other way around? I.e. can chemical tagging identify small subsets of stars, among a vastly greater background sample, that have a common birth origin. If we find stars purely by their clustering in abundance space and subsequently show that they are part of a still spatially coherent cluster, then chemical tagging is
\emph{on.}

\section{Data: \apogee\  giants with [X/H] re-derived from \thecannon }

Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg < 3.9$) from \apogee\ \acronym{DR12} (\cite{dr12}), alll those not flagged as bad in the \apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions. We re-analyze these
\emph{spectra} using \thecannon\ , because this deliver stellar abundance labels of higher precision, especially for stars of SNR$\le 150$.

The details about how we select, reduce, and analyze the \apogee\ data are given in full detail in the companion papers \citep{casey16, ness16}, and we only summarize briefly here.
We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
referenced to H) form the \apogee\ \acronym{DR12} spectra. 
In the \thecannon 's  training step we trained on the
ARC:XXXXX red-giant stars that have spectral SNR$\ge 200$.
For this small fraction of the sample with the highest SNR objects the labels provided by \aspcap provide consistent, low-scatter,
sensible abundance-space measurements. The training step fixes a spectral model that predics the normalized spectrum as a quadratic function of all labels. In the second stage---the test step---each unlabeled star is used to establish a single-star likelihood function for its labels, holding the model
coefficients fixed.
\thecannon\ finds the labels that minimize the single-star likelihood
function for each test-set star.
This optimization is not convex, but it is embarrassingly parallel,
and therefore fast. The test
step takes less than an hour on a small research cluster in Cambridge.

% We start by performing our own continuum normalization on individual-exposure spectra and co-add them ourselves.
% This is required, because the performance of \thecannon\ depends
% critically on the methods used for spectral normalization.  It does
% not require accurate continuum estimation, but it requires that the
% normalization vector be obtained by a \emph{linear operation on the
%   data}. The linear operation we use is a smooth-function fit to pixels
% identified as continuum in our previous work \citep{thecannon}.

% For the training data labels (used to train \thecannon; see below), we
% use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
% 15-element abundances, but only for the subset of ARC:XXXXX red-giant
% stars that have spectral signal-to-noise ratio measurements greater
% than 200.
% This is a small fraction of the total sample, but these highest
% signal-to-noise stars are observed to obtain consistent, low-scatter,
% sensible abundance-space measurements.

% As described in the original paper \citep{thecannon} and the companion
% papers \citep{casey16, ness16}, \thecannon\ is operated in two stages.
% In the first stage---the training step---the training set of stars with
% measured spectra and known labels (known parameters and abundances) is
% used to establish a likelihood function for a set of coefficients that
% parameterize a smooth, simple prediction of spectral pixels given
% labels.
% \thecannon\ finds the set of coefficients that optimizes the
% training-step likelihood function.
% In the implementation used in this \documentname, the spectral
% prediction is quadratic in labels (including all cross-terms to second
% order), and there are 17 labels ($\teff$, $\logg$, and 15 abundances
% referenced to H) and 8500 spectral pixels, so there are on the order
% of 1.4 million coefficients set by a convex optimization in the
% training step.

% In the second stage---the test step---each unlabeled star is used to
% establish a single-star likelihood function for labels, holding the
% coefficients fixed.
% \thecannon\ finds the labels that minimize the single-star likelihood
% function for each test-set star.
% This optimization is not convex, but it is embarrassingly parallel.
% For the results shown here, the training step takes hours and the test
% step takes less than an hour on a small research cluster in Cambridge.
